Fast Fujisaki-Okamoto transformation using encrypt-then-mac and applications to Kyber

### 2024-07-21 etm-transform/manuscript/main.tex
Security analysis

\begin{figure}[H]
    \center
    \begin{minipage}{0.29\textwidth}
        \begin{algorithm}[H]
            \caption{$\keygen_\etm$}\label{alg:etm-keygen}
            \begin{algorithmic}[1]
                \State $(\pk, \sk_\pke) \leftsample \keygen(1^\lambda)$
                \State $z \leftsample \mathcal{M}_\pke$
                \State $\sk \leftarrow (\sk_\pke, z)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \begin{algorithm}[H]
            \caption{$\encap_\etm(\pk)$}\label{alg:etm-encap}
            \begin{algorithmic}[1]
                \State $m \leftsample \mathcal{M}_\pke$
                \State $k_\mac \leftarrow G(\pk, m)$
                \State $c_\pke \leftsample \encrypt(\pk, m)$
                \State $t \leftarrow \mac(k, c_\pke)$
                \State $K \leftarrow H(\pk, m, c_\pke)$
                \State $c \leftarrow (c_\pke, K)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.37\textwidth}
        \begin{algorithm}[H]
            \caption{$\decap_\etm(\sk, c)$}\label{alg:etm-decap}
            \begin{algorithmic}[1]
                \State $(c_\pke, t) \leftarrow c$
                \State $(\sk_\pke, z) \leftarrow \sk$
                \State $\hat{m} \leftarrow \decrypt(\sk_\pke, c_\pke)$
                \State $\hat{k}_\pk \leftarrow G(\pk, \hat{m})$
                \If{$\mac(\hat{k}, c_\pke) \neq t$}
                    \State \Return $H(\pk, z, c_\pke)$
                \EndIf
                \State \Return $H(\pk, \hat{m}, c_\pke)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{$\kem_\etm$ routines}\label{fig:etm-routines} 
\end{figure}

@@

\section{Application to Kyber}
\subsection{``encrypt-then-MAC'' is not secure with Kyber}
CRYSTALS-Kyber \cite{bos2018crystals}\cite{avanzi2019crystals} and ML-KEM \cite{key2023mechanism} are \monospace{IND-CCA2} secure key encapsulation mechanisms whose security depends on the hardness of the Module Learning with Error (MLWE) problem. For the construction of the \monospace{IND-CCA2} secure KEM, Kyber first constructs an \monospace{IND-CPA} public-key encryption scheme (which we will call \monospace{CPAPKE}), then applies the modular Fujisaki-Okamoto transformation \cite{hofheinz2017modular} to construct the key encapsulation mechanism (which we will call \monospace{CCAKEM}). Specifically, Kyber's round-3 submission uses the $U^{\not\bot}$ transformation, while ML-KEM uses the $U^{\not\bot}_m$ transformation. The routines of \monospace{CPAPKE} can be found in Algorithm 4, 5, 6 in \cite{avanzi2019crystals} and are largely identical between Kyber and ML-KEM.

\emph{\monospace{CPAPKE} is not \monospace{OW-PCA} secure}: What's important to know about \monospace{CPAPKE} is that it is not secure against plaintext-checking attack.

>>>>

\begin{figure}[H]
    \center
    \begin{algorithm}[H]
        \caption{$\keygen_\etm$}\label{alg:etm-keygen}
        \begin{algorithmic}[1]
            \State $(\pk, \sk_\pke) \leftsample \keygen(1^\lambda)$
            \State $z \leftsample \mathcal{M}_\pke$
            \State $\sk \leftarrow (\sk_\pke, z)$
            \State \Return $(\pk, \sk)$
        \end{algorithmic}
    \end{algorithm}
    \begin{algorithm}[H]
        \caption{$\encap_\etm(\pk)$}\label{alg:etm-encap}
        \begin{algorithmic}[1]
            \State $m \leftsample \mathcal{M}_\pke$
            \State $k_\mac \leftarrow G(\pk, m)$
            \State $c_\pke \leftsample \encrypt(\pk, m)$
            \State $t \leftarrow \mac(k, c_\pke)$
            \State $K \leftarrow H(\pk, m, c_\pke)$
            \State $c \leftarrow (c_\pke, K)$
            \State \Return $(c, K)$
        \end{algorithmic}
    \end{algorithm}
    \begin{algorithm}[H]
        \caption{$\decap_\etm(\sk, c)$}\label{alg:etm-decap}
        \begin{algorithmic}[1]
            \State $(c_\pke, t) \leftarrow c$
            \State $(\sk_\pke, z) \leftarrow \sk$
            \State $\hat{m} \leftarrow \decrypt(\sk_\pke, c_\pke)$
            \State $\hat{k}_\pk \leftarrow G(\pk, \hat{m})$
            \If{$\mac(\hat{k}, c_\pke) \neq t$}
                \State \Return $H(\pk, z, c_\pke)$
            \EndIf
            \State \Return $H(\pk, \hat{m}, c_\pke)$
        \end{algorithmic}
    \end{algorithm}
    \caption{$\kem_\etm$ routines}\label{fig:etm-routines} 
\end{figure}

@@

\section{Application to Kyber}

CRYSTALS-Kyber \cite{bos2018crystals}\cite{avanzi2019crystals} and ML-KEM \cite{key2023mechanism} are \monospace{IND-CCA2} secure key encapsulation mechanisms whose security depends on the hardness of the Module Learning with Error (MLWE) problem. For the construction of the \monospace{IND-CCA2} secure KEM, Kyber first constructs an \monospace{IND-CPA} public-key encryption scheme (which we will call \monospace{CPAPKE}), then applies the modular Fujisaki-Okamoto transformation \cite{hofheinz2017modular} to construct the key encapsulation mechanism (which we will call \monospace{CCAKEM}). Specifically, Kyber's round-3 submission uses the $U^{\not\bot}$ transformation, while ML-KEM uses the $U^{\not\bot}_m$ transformation. The routines of \monospace{CPAPKE} can be found in Algorithm 4, 5, 6 in \cite{avanzi2019crystals} and are largely identical between Kyber and ML-KEM.

The routines of \monospace{CCAKEM} can be found in Algorithm 7, 8, 9 in \cite{avanzi2019crystals}. We recap them here:

\begin{algorithm}[H]
    \caption{\monospace{Kyber.CCAKEM.KeyGen()}}
    \begin{algorithmic}[1]
        \State $z \leftsample \mathcal{B}^{32}$
            \Comment{Randomly sample 32 bytes (256 bits)}
        \State $(\pk, \sk^\prime) \leftsample \monospace{Kyber.CPAPKE.KeyGen()}$
        \State $\sk = (\sk^\prime, \pk, H(\pk), z)$
            \Comment{H is instantiated with SHA3-256}
        \State \Return $(\pk, \sk)$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{\monospace{Kyber.CCAKEM.Encap(\pk)}}
    \begin{algorithmic}[1]
        \State $m \leftsample \mathcal{B}^{32}$
        \State $m \leftarrow H(m)$
            \Comment{Do not output system RNG directly}
        \State $(\bar{K}, r) \leftarrow G(m \Vert H(\pk))$
            \Comment{G is instantiated with SHA3-512}
        \State $c \leftarrow \monospace{Kyber.CPAPKE.Enc(\pk, m, r)}$
            \Comment{Because $r$ is set, \monospace{CPAPKE} is deterministic}
        \State $K \leftarrow \monospace{KDF}(\bar{K} \Vert H(c))$
            \Comment{\monospace{KDF} is instantiated with Shake256}
        \State \Return $(c, K)$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{\monospace{Kyber.CCAKEM.Decap(\sk, c)}}
    \begin{algorithmic}[1]
        \Require Secret key $\sk = (\sk^\prime, \pk, H(\pk), z)$
        \Require \monospace{Kyber.CPAPKE} Ciphertext $c$
        \State $(\sk^\prime, \pk, h, z) \leftarrow \sk$
            \Comment{Unpack the secret key; $h$ is the hash of $\pk$}
        \State $m^\prime \leftarrow \monospace{Kyber.CPAPKE.Dec}(\sk^\prime, c)$
        \State $(\overline{K}^\prime, r^\prime) \leftarrow G(m^\prime \Vert h)$
        \State $c^\prime = \monospace{Kyber.CPAPKE.Enc}(\pk, m^\prime, r^\prime)$
        \If{$c = c^\prime$}
            \State $K \leftarrow \monospace{KDF}(\bar{K}^\prime \Vert H(c))$
        \Else
            \State $K \leftarrow \monospace{KDF}(z \Vert H(c))$
        \EndIf
        \State \Return $K$
    \end{algorithmic}
\end{algorithm}

\subsection{``encrypt-then-MAC'' is not secure with Kyber}
\begin{lemma}
    There exists efficient \monospace{OW-PCA} adversary against \monospace{CPAPKE}
\end{lemma}
